---
title: "Covariates in Population and Catchability Components"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Covariates in Population and Catchability Components}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

`jointCPUE()` supports covariates in two distinct model components:

- `formula_catchability` describes the observation or catchability process.
  Its effects are fitted to observations but excluded from the standardized
  abundance index.
- `formula_population` describes population density. Its effects are fitted to
  observations and projected over the extrapolation grid, so they enter the
  standardized index.

This distinction should be made from the biological meaning of a covariate,
not from which specification gives a better fit. For example, fishing depth
may describe gear availability and belong in catchability, whereas an
environmental variable used to describe habitat may belong in population
density.

# Supported formula terms

Both arguments use one-sided formulas. The following terms can be combined:

| Formula term | Required input type | Interpretation |
|---|---|---|
| `x` | numeric or integer | linear fixed effect |
| `x` | factor or character | fixed factor effect with sum-to-zero coding |
| `s(x)` | numeric | penalized smooth effect |
| `f(x)` | values defining groups | independent random factor effects with an estimated SD |

For example:

```{r}
fit <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  formula_catchability = ~ s(depth) + f(season),
  formula_population = ~ s(sst) + chl
)
```

Here, `depth` and `sst` have smooth effects, `chl` has a linear fixed effect,
and the levels of `season` have random effects.

## Numeric versus factor variables

For a simple fixed term such as `~ season`, the R column type determines the
effect:

```{r}
## Integer values give a linear fixed effect.
data_utm$season <- c(1L, 2L, 3L, 4L)[data_utm$season_id]

## A factor gives a sum-to-zero fixed factor effect.
data_utm$season <- factor(
  data_utm$season,
  levels = 1:4,
  labels = c("winter", "spring", "summer", "autumn")
)
```

The model cannot infer that integer values are category labels. Always convert
numeric category codes to `factor` before using a simple fixed term. Character
columns are also treated as factors, but an explicit factor is preferable
because it records the intended level order.

A fixed factor with \(K\) observed levels is represented by \(K-1\)
coefficients, with the final level implied so that all \(K\) level effects sum
to zero. In contrast, `f(season)` creates \(K\) random effects with a common
estimated standard deviation and does not impose a hard sum-to-zero constraint.

# Catchability covariates

Catchability covariates require values only at the observations:

```{r}
data_utm$season <- factor(data_utm$season)

fit_catchability <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  formula_catchability = ~ s(depth) + season,
  pop_spatial = "on",
  pop_spatiotemporal = "on",
  q_diffs_system = "on"
)
```

In this model, `s(depth)` is a smooth observation-process effect and `season`
is a sum-to-zero fixed factor effect. Neither effect is included when the
standardized index is calculated.

Use `f(season)` instead when seasons should be partially pooled through a
random-effect distribution:

```{r}
fit_random_season <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  formula_catchability = ~ s(depth) + f(season)
)
```

# Population covariates and projection data

Population covariates enter the index and therefore need values at every
extrapolation-grid cell and modeled time. Static covariates can be supplied
with one row per grid cell:

```{r}
population_projection <- data.frame(
  utm_x_scale = extrapolation_grid$utm_x_scale,
  utm_y_scale = extrapolation_grid$utm_y_scale,
  sst_climatology = extrapolation_grid$sst_climatology,
  chl_climatology = extrapolation_grid$chl_climatology
)

fit_static_environment <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  extrapolation_grid = extrapolation_grid,
  formula_population = ~ s(sst_climatology) + s(chl_climatology),
  projection_data = population_projection
)
```

The `utm_x_scale` and `utm_y_scale` values in `projection_data` must match the
extrapolation grid. Rows may be supplied in any order; they are aligned by
these coordinates.

For time-varying covariates, supply one row for every grid cell and time. The
`tid` values must use the same 0-based coding as `data_utm$tid`:

```{r}
population_projection_time <- expand.grid(
  grid_row = seq_len(nrow(extrapolation_grid)),
  tid = sort(unique(data_utm$tid))
)

population_projection_time$utm_x_scale <-
  extrapolation_grid$utm_x_scale[population_projection_time$grid_row]
population_projection_time$utm_y_scale <-
  extrapolation_grid$utm_y_scale[population_projection_time$grid_row]

## Join or assign one environmental value per grid cell-time combination.
population_projection_time$sst <- projected_sst
population_projection_time$grid_row <- NULL

fit_time_environment <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  extrapolation_grid = extrapolation_grid,
  formula_population = ~ s(sst),
  projection_data = population_projection_time
)
```

If each population covariate has one unambiguous value within every
extrapolation cell, `projection_data` may be omitted and the values are inferred
from `data_utm`. Supplying it explicitly is recommended for time-varying
environmental variables and whenever observation locations do not define the
desired projection values.

# Yearly month effects

The existing `month_diffs` interface remains available for yearly models:

```{r}
fit_yearly <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  index = "yearly",
  month_col = "month",
  month_diffs = "on",
  formula_catchability = ~ s(depth)
)
```

This special month effect is a sum-to-zero fixed effect that enters only the
observation model. It is excluded from the yearly index. Do not also include
`month` in `formula_catchability` or `formula_population` when
`month_diffs = "on"`; `jointCPUE()` stops with an error if the effect is
specified twice.

# Practical model-building guidance

Add covariates in biologically motivated groups and compare convergence,
residual patterns, and index sensitivity. Smooth and random-factor terms add
variance parameters, so weakly informed or highly correlated terms can make
optimization harder. The population/catchability assignment also changes the
estimand: moving a term between formulas changes whether its effect is retained
in the standardized index.
