---
title: "Preparing NPFC Neon Flying Squid Data"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Preparing NPFC Neon Flying Squid Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

This vignette documents the data-preparation workflow I used for the NFS
examples in `jointCPUE`.

The workflow covers:

- importing member-specific source files
- standardizing columns into a common format
- splitting Taiwan records by gear configuration
- creating 0-based yearly and monthly indices
- plotting fleet coverage in space and time
- splitting the data at `170E`

Replace the file paths and column positions as needed for your local data
files. The column selections shown here are the ones I used for the current
NFS files.

# Load packages

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(maps)
library(tidyr)
```

# Read raw files

```{r}
jpn <- read_excel(".../20251125Data sharing template NFS Japan (by Month and Grid)_2017-2024.xlsx")
twn <- read_excel(".../1x1-degree catch and effort data NFS_CT.xlsx")
rus <- read_excel(".../NFS_RUS_NW_ed1.xlsx")
kor <- read_excel(".../Data sharing templete NFS up to 2025_KOR.xlsx")
chn <- read.csv(".../CE_NFS_China_2005-2024_0.csv")
```

# Data notes

- `JPN`: 2017-2024; machine jigging; January-February and April-December
- `TWN`: mixed handline and machine-jigging records
- `RUS`: drift-net; effort in vessel-days
- `KOR`: gear type not clearly distinguished

# Split Taiwan records by effort configuration

```{r}
twn <- twn %>%
  mutate(
    eff1 = suppressWarnings(as.numeric(`Effort number 1`)),
    eff2 = suppressWarnings(as.numeric(`Effort number 2`))
  )

twn_hand <- twn %>% filter(!is.na(eff1), eff1 > 0, is.na(eff2) | eff2 == 0)
twn_mac  <- twn %>% filter((is.na(eff1) | eff1 == 0), !is.na(eff2), eff2 > 0)
twn_both <- twn %>% filter(!is.na(eff1), eff1 > 0, !is.na(eff2), eff2 > 0)
twn_no   <- twn %>% filter((is.na(eff1) | eff1 == 0), is.na(eff2) | eff2 == 0)
```

# Remove known unusable records

```{r}
jpn <- jpn %>% filter(Area != "NW")
```

# Standardize member files

```{r}
parse_month <- function(x) {
  x <- trimws(as.character(x))

  dplyr::case_when(
    x %in% c("1", "01", "January", "JAN", "Jan") ~ 1L,
    x %in% c("2", "02", "February", "FEB", "Feb") ~ 2L,
    x %in% c("3", "03", "March", "MAR", "Mar") ~ 3L,
    x %in% c("4", "04", "April", "APR", "Apr") ~ 4L,
    x %in% c("5", "05", "May", "MAY") ~ 5L,
    x %in% c("6", "06", "June", "JUN", "Jun") ~ 6L,
    x %in% c("7", "07", "July", "JUL", "Jul") ~ 7L,
    x %in% c("8", "08", "August", "AUG", "Aug") ~ 8L,
    x %in% c("9", "09", "September", "SEP", "Sep") ~ 9L,
    x %in% c("10", "October", "OCT", "Oct") ~ 10L,
    x %in% c("11", "November", "NOV", "Nov") ~ 11L,
    x %in% c("12", "December", "DEC", "Dec") ~ 12L,
    TRUE ~ NA_integer_
  )
}

mk_dd <- function(dat, cols, fleet) {
  out <- dat[, cols]
  colnames(out) <- c(
    "Member", "Year", "Month", "Latitude", "Longitude", "Longitude_EW",
    "Effort_type", "Effort", "Catch"
  )

  out %>%
    mutate(
      Member = as.character(Member),
      Year = as.integer(Year),
      Month = parse_month(Month),
      Latitude = as.numeric(Latitude),
      Longitude = as.numeric(Longitude),
      Longitude_EW = as.character(Longitude_EW),
      Effort_type = as.character(Effort_type),
      Effort = as.numeric(Effort),
      Catch = as.numeric(Catch),
      Fleet = fleet
    )
}
```

```{r}
dd <- bind_rows(
  mk_dd(chn,      c(1:3, 5:7, 10,  9, 12), "CHN"),
  mk_dd(jpn,      c(1:3, 6:8, 13, 12, 14), "JPN"),
  mk_dd(twn_hand, c(1:3, 5:7, 11, 12, 15), "TWN_hand"),
  mk_dd(twn_mac,  c(1:3, 5:7, 13:15),      "TWN_machine"),
  mk_dd(kor,      c(1:3, 5:7, 11:13),      "KOR"),
  mk_dd(rus,      c(1:3, 5:7, 11:13),      "RUS")
)
```

# Create analysis fields

```{r}
dd <- dd %>%
  mutate(
    X = if_else(Longitude_EW == "W", 360 - Longitude, Longitude),
    Y = Latitude,
    Fleet_code = case_when(
      Fleet == "CHN" ~ 0L,
      Fleet == "JPN" ~ 1L,
      Fleet == "KOR" ~ 2L,
      Fleet == "RUS" ~ 3L,
      Fleet == "TWN_hand" ~ 4L,
      Fleet == "TWN_machine" ~ 5L,
      TRUE ~ NA_integer_
    ),
    Presence = if_else(Catch > 0, 1L, 0L),
    CPUE = Catch / Effort
  )
```

# Build 0-based observed-time indices

These indices skip missing years and missing year-month combinations.

```{r}
year_lookup <- dd %>%
  distinct(Year) %>%
  arrange(Year) %>%
  mutate(t_year = row_number() - 1L)

month_lookup <- dd %>%
  distinct(Year, Month) %>%
  arrange(Year, Month) %>%
  mutate(t_month = row_number() - 1L)

dd <- dd %>%
  select(-any_of(c("t_year", "t_month"))) %>%
  left_join(year_lookup, by = "Year") %>%
  left_join(month_lookup, by = c("Year", "Month"))
```

# Plot fleet coverage

```{r}
world <- map_data("world") %>%
  mutate(long_360 = ifelse(long < 0, long + 360, long))

x_rng <- range(dd$X, na.rm = TRUE)
y_rng <- range(dd$Y, na.rm = TRUE)

plot_spatial <- ggplot() +
  geom_polygon(
    data = world %>%
      filter(
        long_360 >= x_rng[1] - 5,
        long_360 <= x_rng[2] + 5,
        lat >= y_rng[1] - 3,
        lat <= y_rng[2] + 3
      ),
    aes(x = long_360, y = lat, group = group),
    fill = "grey85",
    color = "grey50",
    linewidth = 0.2
  ) +
  geom_point(
    data = dd,
    aes(x = X, y = Y, color = Fleet, shape = Fleet),
    size = 2.5,
    alpha = 0.8
  ) +
  coord_quickmap(
    xlim = c(135, 235),
    ylim = c(15, 52),
    expand = FALSE
  ) +
  scale_x_continuous(
    labels = function(x) ifelse(x <= 180, paste0(x, "E"), paste0(360 - x, "W"))
  ) +
  labs(x = "Longitude", y = "Latitude") +
  theme_bw()
```

```{r}
plot_temporal_coverage <- function(dat, month_lookup) {
  plot_dat <- tidyr::expand_grid(
    Fleet = unique(dat$Fleet),
    t_month = month_lookup$t_month
  ) %>%
    left_join(month_lookup, by = "t_month") %>%
    left_join(
      dat %>%
        distinct(Fleet, Year, Month) %>%
        mutate(has_data = 1L),
      by = c("Fleet", "Year", "Month")
    ) %>%
    mutate(fill_group = if_else(is.na(has_data), "No data", Fleet))

  ggplot(plot_dat, aes(x = t_month, y = Fleet, fill = fill_group)) +
    geom_tile(width = 0.85, height = 0.75, color = NA) +
    scale_fill_manual(
      values = c(
        "No data" = "grey92",
        "CHN" = "#C94C4C",
        "JPN" = "#E76F51",
        "KOR" = "#457B9D",
        "RUS" = "#8D99AE",
        "TWN_hand" = "#F4A261",
        "TWN_machine" = "#2A9D8F"
      )
    ) +
    scale_x_continuous(
      breaks = month_lookup$t_month[month_lookup$Month == 1],
      labels = month_lookup$Year[month_lookup$Month == 1]
    ) +
    labs(x = "Year-Month", y = NULL, fill = NULL) +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}

plot_temporal <- plot_temporal_coverage(dd, month_lookup)
```

# Split the data at 170E

```{r}
dd_left_170 <- dd %>% filter(X < 170)
dd_right_170 <- dd %>% filter(X >= 170)
```

```{r}
plot_temporal_left <- plot_temporal_coverage(dd_left_170, month_lookup)
plot_temporal_right <- plot_temporal_coverage(dd_right_170, month_lookup)
```

# Save processed outputs

```{r}
save(
  dd_left_170,
  dd_right_170,
  year_lookup,
  month_lookup,
  file = "NFS_jointCPUE_Data_split170E.RData"
)
```

# Notes

- `Fleet_code` is 0-based, with `CHN` used as the reference fleet.
- `t_year` and `t_month` are also 0-based, and only observed time points are
  indexed.
- If you need full calendar time indices with gaps retained, build the lookup
  tables from a complete year or year-month sequence instead of observed values
  only.
