---
title: "NFS Right of 170E: Monthly CPUE Workflow"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{NFS Right of 170E: Monthly CPUE Workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

This vignette documents the monthly CPUE workflow for the NFS data east of
`170E`.

The workflow is:

- loading the processed `170E` split dataset
- fitting a sequence of increasingly complex monthly models
- computing marginal AIC values
- extracting standardized indices
- checking fitted values and residual patterns
- plotting fleet catchability differences when enabled

The code below provides a reproducible workflow. File paths and model settings
should be modified for local use.

The monthly comparison in this vignette uses an iid spatiotemporal field. After
the data are split and indexed on observed months, the monthly series is no
longer a complete calendar sequence, so an RW specification is not the default
comparison.

# Load package and processed data

```{r}
library(jointCPUE)

load("NFS_jointCPUE_Data_split170E.RData")

## Use the region east of 170E
dd <- dd_right_170
```

# Prepare model input

```{r}
data_input <- data.frame(
  cpue = dd$CPUE + 1e-05,
  lon = dd$X,
  lat = dd$Y,
  tid = as.numeric(as.factor(dd$t_month)) - 1,
  fleetid = dd$Fleet_code
)
```

```{r}
utm <- make_utm(data_input, utm_zone = NULL, coord_scale = "auto")
data_utm <- utm$data_utm
```

```{r}
mesh <- make_mesh(
  data_utm,
  xy_cols = c("utm_x_scale", "utm_y_scale"),
  type = "cutoff",
  cutoff = 0.5
)

plot(mesh)
```

```{r}
ncores <- 32
```

# Model 1: Fleet systematic differences only

```{r}
fit_1 <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  pop_spatial = "off",
  pop_spatiotemporal = "off",
  q_diffs_system = "on",
  q_diffs_time = "off",
  q_diffs_spatial = "off",
  ncores = ncores
)
```

```{r}
check_convergence(fit_1)
```

```{r}
calc_marginal_aic(fit_1)

index_1 <- get_index(fit_1)
plot_index(index_1, time_values = paste(month_lookup_right_170$Year, month_lookup_right_170$Month, sep = "-"))
```

```{r}
pred_1 <- get_predicted(
  fit_1,
  data = data_utm,
  drop_floor = TRUE,
  floor_value = 1e-5
)

plots_1 <- plot_residuals(
  pred_1,
  observed_col = "cpue",
  x_col = "utm_x_scale",
  y_col = "utm_y_scale",
  residual = "log",
  bins = 40
)

plots_1$observed_predicted
plots_1$spatial_residual
```

# Model 2: Add population spatial field

```{r}
fit_2 <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  pop_spatial = "on",
  pop_spatiotemporal = "off",
  q_diffs_system = "on",
  q_diffs_time = "off",
  q_diffs_spatial = "off",
  ncores = ncores
)
```

```{r}
check_convergence(fit_2)
```

```{r}
calc_marginal_aic(fit_2)

index_2 <- get_index(fit_2)
plot_index(index_2, time_values = paste(month_lookup_right_170$Year, month_lookup_right_170$Month, sep = "-"))
```

```{r}
pred_2 <- get_predicted(
  fit_2,
  data = data_utm,
  drop_floor = TRUE,
  floor_value = 1e-5
)

plots_2 <- plot_residuals(
  pred_2,
  observed_col = "cpue",
  x_col = "utm_x_scale",
  y_col = "utm_y_scale",
  residual = "log",
  bins = 40
)

plots_2$observed_predicted
plots_2$spatial_residual
```

# Model 3: Add iid spatiotemporal field

```{r}
fit_3 <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  pop_spatial = "on",
  pop_spatiotemporal = "on",
  pop_spatiotemporal_type = "iid",
  q_diffs_system = "on",
  q_diffs_time = "off",
  q_diffs_spatial = "off",
  ncores = ncores
)
```

```{r}
check_convergence(fit_3)
```

```{r}
calc_marginal_aic(fit_3)

index_3 <- get_index(fit_3)
plot_index(index_3, time_values = paste(month_lookup_right_170$Year, month_lookup_right_170$Month, sep = "-"))
```

```{r}
pred_3 <- get_predicted(
  fit_3,
  data = data_utm,
  drop_floor = TRUE,
  floor_value = 1e-5
)

plots_3 <- plot_residuals(
  pred_3,
  observed_col = "cpue",
  x_col = "utm_x_scale",
  y_col = "utm_y_scale",
  residual = "log",
  bins = 40
)

plots_3$observed_predicted
plots_3$spatial_residual
```

# Model 4: Add fleet-specific spatial differences

```{r}
fit_4 <- jointCPUE(
  data_utm = data_utm,
  mesh = mesh,
  pop_spatial = "on",
  pop_spatiotemporal = "on",
  pop_spatiotemporal_type = "iid",
  q_diffs_system = "on",
  q_diffs_time = "off",
  q_diffs_spatial = "on",
  ncores = ncores
)
```

```{r}
check_convergence(fit_4)
```

```{r}
calc_marginal_aic(fit_4)

index_4 <- get_index(fit_4)
plot_index(index_4, time_values = paste(month_lookup_right_170$Year, month_lookup_right_170$Month, sep = "-"))
```

```{r}
pred_4 <- get_predicted(
  fit_4,
  data = data_utm,
  drop_floor = TRUE,
  floor_value = 1e-5
)

plots_4 <- plot_residuals(
  pred_4,
  observed_col = "cpue",
  x_col = "utm_x_scale",
  y_col = "utm_y_scale",
  residual = "log",
  bins = 40
)

plots_4$observed_predicted
plots_4$spatial_residual
```

# Plot fleet catchability differences

```{r}
fleet_labels <- c("CHN", "JPN", "KOR", "RUS", "TWN_hand", "TWN_machine")
```

```{r}
plot_q_diffs_system(
  fit_4,
  fleet_labels = fleet_labels
)
```

```{r}
plot_q_diffs_spatial(
  fit_4,
  fleet_labels = fleet_labels
)
```

If you fit a model with `q_diffs_time = "on"`, monthly fleet-specific temporal
deviations can be plotted as follows:

```{r}
plot_q_diffs_time(
  fit_time,
  fleet_labels = fleet_labels,
  time_values = paste(month_lookup_right_170$Year, month_lookup_right_170$Month, sep = "-")
)
```

# Notes

- The right-of-`170E` analysis is based on `dd_right_170`.
- The response is modeled as `CPUE + 1e-5` to avoid zero values in the
  lognormal likelihood.
- Residual plots are shown after removing the artificial floor value with
  `drop_floor = TRUE`.
- `plot_index()` is called with `time_values` set to year-month labels from
  `month_lookup_right_170` so that the right-of-`170E` monthly index is shown on
  the observed year-month scale.
